Are environmental and geographic variables effective surrogates for genetic variation in conservation planning?

### 2016-04-01 article/article.Rmd
+ edits to article

## Introduction
<!--  main text -->
The act of setting aside areas for preservation spearheads modern conservation efforts [@r440]. Protected area networks buffer species from anthropogenic impacts and provide places for populations of species to persist. Intra-specific genetic variation underpins species persistence, and there has been increasing interest in understanding how to design protected area networks in a way that maximizes this variation [@r67; @r3; @r61]. Broadly speaking, genetic variation can be classified as either adaptive or neutral [reviewed in @r48]. Adaptive genetic variation is associated with loci that affect fitness. By preserving existing patterns of adaptive variation, protected areas can ensure that populations with particularly beneficial adaptations are not lost [@r3]. In contrast, neutral genetic variation is associated with loci that do not affect fitness. Neutral variation often reflects the evolutionary history of different populations. By preserving existing patterns of neutral variation, protected areas can secure different genetic lineages and avoid the adverse effects of low genetic diversity [eg. inbreeding depression; @r67]. However, genetic data requires considerable investment to obtain and analyze. As a consequence, conservation planners do not typically use genetic data to inform prioritizations [@r61]. Thus prioritizations may fail to capture species' intra-specific genetic variation, and ultimately fail to deliver long-term conservation outcomes.

Recently, conservation scientists have been investigating surrogates for genetic variation to generate better prioritizations without the overheads associated with genetic data. Since adaptive genetic variation is ultimately driven by extrinsic selection pressures, environmental variables have been used to guide reserve selection [eg. habitat conditions; @r247; @r2]. However, their effectiveness as surrogates in conservation planning remains unverified. On the other hand, neutral genetic variation arises from a reduction in gene flow between individuals. Surrogates for neutral variation are based on variables that predict the level of connectivity between planning units. For instance, the isolation by distance theory [IBD; @r259] predicts that populations further apart will have less connectivity, exchange less genetic material, and in turn have less genetic variation in common. Based on this theory, geographic distance-based surrogates have been found to improve prioritizations in an island network [@r200]. However this remains untested in spatially contiguous systems where connectivity is complicated by additional factors [eg. terrain; eg. @r474]--the very systems commonly featured in real-world conservation planning exercises [eg. @r197]. Before conservation planners can reliably use surrogates for genetic data in conservation planning, the effectiveness of these surrogates needs to be verified. Otherwise, conservation planners may waste precious resources on protecting places that provide little benefit to biodiversity.

Here we investigate the effectiveness of environmental and geographic surrogates for adaptive and neutral genetic variation in the context of conservation planning. First, we examine the relationship between surrogates and genetic variation, and verify whether these variables are effective surrogates. Second, we determine if the inclusion of surrogates in various conservation planning scenarios can improve the proportion of genetic variation they secure. If these surrogates can improve the intra-specific genetic variation captured in a prioritization, then conservation planners may use these surrogates when genetic data is not available, and in turn deliver more effective protected areas.

@@

### Prioritizations
We used the unreliable representative and adequate prioritization (URAP) formulation of the reserve selection problem [@r465]. This formulation can identify the minimum number of planning units required to preserve both a proportion of the species' range (using amount-based targets) and a proportion of the species' intra-specific variation [using space-based targets and attribute spaces; _sensu_ @r465]. We treated the environmental and geographic surrogate variables as separate attribute spaces. We also treated the adaptive (if detected) and neutral variables as separate attribute spaces for each species. We set the demand points for each species and each attribute space using the values associated with the planning units. For every prioritization we generated, we calculated the proportion of the adaptive (if detected) and neutral genetic variation that every prioritization secured to assess its performance. Prioritizations associated with negative values--because they secured only a very small amount of genetic variation--were replaced with zeros to facilitate statistical analysis. We solved all reserve selection problems to within `r (gurobi.params.LST[[MODE]]$MIPGap*100)` % of optimality using the $\texttt{rapr}$ R package [@r465] and $\texttt{Gurobi}$ [@r463].

To determine if the environmental and geographic variables are effective surrogates for adaptive and neutral genetic variation, we generated a collection of single-species prioritizations. For each species, we first generated two sets of prioritizations using increasing environmental and geographic-based targets (separately). Prioritizations were generated using the following surrogate-based targets: `r paste(rapr.params.LST[[MODE]]$surrogate.target,collapse=', ')`. We then create a set of null expectations with which to compare the surrogate-based prioritizations. These null expectations represented possible solutions obtained using conventional conservation planning methods that just use amount (eg. Marxan). For each species, for each target, we generated `r ` prioritizations by randomly selecting planning units occupied by the species. The number of randomly selected planning units was set to equal the number of planning units in the optimal prioritizations.

To determine if the use of surrogate-based targets can actually improve prioritizations in real-world conservation planning scenarios, we investigated the effectiveness of surrogate-based targets using three scenarios with increasing levels of complexity. To assess the benefits of surrogate-based targets under the simplest of circumstances, our first scenario involved generating single-species prioritizations for each species. However, most large-scale prioritizations involve a comprehensive set of species. By forcing solutions to secure individuals in multiple communities, this may result in prioritizations that secure a greater proportion of intra-specific genetic variation for each species. To test for this effect, our second scenario involved generating multi-species prioritizations. Finally, most large-scale prioritizations also explicitly consider opportunity cost to ensure that prioritizations are cost-effective. Our third scenario involved generating multi-species prioritizations that satisfied targets for minimal opportunity cost.

For each scenario, we generated three types of prioritizations using (1) only amount-based targets, (2) both amount- and surrogate-based targets, and (3) both amount- and genetic-based targets. The purely amount-based prioritizations served as our baseline--representing the prioritizations typically generated in conservation planning exercises. We used `r (rapr.params.LST[[MODE]]$scenario.analysis$amount.target*100)` % amount-based targets in all prioritizations to ensure an adequate proportion of habitat was secured for each species. Based on the previous analysis, we set surrogate-based targets as `r (rapr.params.LST[[MODE]]$scenario.analysis$surrogate.target*100)` % and genetic-based targets as `r (rapr.params.LST[[MODE]]$scenario.analysis$genetic.target*100)` % to ensure that most of the variation was secured for each species. To accomodate variation in the high number of optimal solutions for the single-species amount-based prioritizations, we generated `r (rapr.params.LST[[MODE]]$scenario.analysis$single.species.amount.replicates)` replicate prioritizations.

## Statistical analysis
We investigated the effectiveness of prioritizations generated using surrogate-based targets using generalized linear mixed-effects models [GLMMs; using a logit link and the $\texttt{lme4}$ R package; @r469]. To determine  if the environmental and geographic variables were effective surrogates, we fit a full model to the proportion of genetic variation that the first collection of prioritizations. This full model contained categorical predictor variables to indicate the surrogate-based targets used to generate them, the type of genetic variation (adaptive or neutral), and the interactions between these variables. To determine if surrogate-based targets improved the effectiveness of prioritizations under various scenarios, we fit a full model to the proportion of genetic variation that second collection of prioritizations secured. This second full model contained categorical predictor variables to indicate the scenario (single-species, multi-species, multi-species with cost), the combination of targets used to generate the prioritizations (amount-based targets, amount and surrogate-based targets, amount and genetic-based targets), the type of genetic variation measured (adaptive or neutral), and the interactions between them. All models contained intercept-level random effects to accommodate variation between species. We conducted a post-hoc analysis on each full model using Tukey contrasts with a Bonferroni correction [using the $\texttt{multcomp}$ R package; @r470].

## Results
<!-- Surrogacy analysis -->
The environmental and geographic-based variables were effective surrogates for adaptive and genetic variation (Figure 1). The prioritizations generated using greater surrogate-based targets secured more genetic variation (XXXX). In addition, prioritizations generated using environmental targets often secured more adaptive genetic variation than prioritizations generated using geographic-based targets did neutral genetic variation (XXXX). Only particularly high geographic targets ($\geq 0.95$ %) were able to secure a moderate proportion of neutral variation ($\geq XXXX$ %). Overall, there was a non-linear relationship between the surrogate-based targets used to generate the prioritizations and the proportion of genetic variation they secured.

>>>>

## Introduction
<!--  main text -->
The act of setting aside areas for preservation spearheads modern conservation efforts [@r440]. Protected area networks buffer species from anthropogenic impacts and provide places for populations of species to persist. Intra-specific genetic variation underpins species persistence, and there has been increasing interest in understanding how to design protected area networks in a way that maximizes this variation for minimal cost [@r67; @r3; @r61]. Broadly speaking, genetic variation can be classified as either adaptive or neutral [reviewed in @r48]. Adaptive genetic variation is associated with loci that affect fitness. By preserving existing patterns of adaptive variation, protected areas can ensure that populations with particularly beneficial adaptations are not lost [@r3]. In contrast, neutral genetic variation is associated with loci that do not affect fitness. Neutral variation often reflects the evolutionary history of different populations. By preserving existing patterns of neutral variation, protected areas can secure different genetic lineages and avoid the adverse effects of low genetic diversity [eg. inbreeding depression; @r67]. However, genetic data requires considerable investment to obtain and analyze. As a consequence, conservation planners do not typically use genetic data to inform prioritizations [@r61]. Thus prioritizations may fail to capture species' intra-specific genetic variation, and ultimately fail to deliver long-term conservation outcomes.

Recently, conservation scientists have been investigating surrogates for intra-specific genetic variation to generate prioritizations that secure itra-specific genetic variation without the overheads associated with using it. Since adaptive genetic variation is ultimately driven by extrinsic selection pressures, environmental variables have been used to guide reserve selection [eg. habitat conditions; @r247; @r2]. However, their effectiveness as surrogates in conservation planning remains unverified. On the other hand, neutral genetic variation arises from a reduction in gene flow between individuals. Surrogates for neutral variation are based on variables that predict the level of connectivity between planning units. For instance, the isolation by distance theory [IBD; @r259] predicts that populations further apart will have less connectivity, exchange less genetic material, and in turn have less genetic variation in common. Based on this theory, geographic distance-based surrogates have been found to improve prioritizations in an island network [@r200]. However this remains untested in spatially contiguous systems where connectivity is complicated by additional factors [eg. terrain; eg. @r474]--the very systems commonly featured in real-world conservation planning exercises [eg. @r197]. Before conservation planners can reliably use surrogates for genetic data in conservation planning, the effectiveness of these surrogates needs to be verified. Otherwise, conservation planners may waste precious resources on protecting places that provide little benefit to biodiversity.

Here we investigate the effectiveness of environmental and geographic surrogates for adaptive and neutral genetic variation in the context of conservation planning. Conventionally, prioritizations are generated by setting targets on the total amount of habitat that needs to be preserved for each species (amount-based targets), and finding the combination of planning units that fulfills all the targets for minimal cost [eg. using \texttt{Marxan}; @r22]. First, we aim to verify that environmental and geographic variables are effective surrogates for adaptive and neutral variation (respectively). We expect that prioritizations generated using surrogate-based targets will secure more genetic variation than prioritizations generated using only amount-based targets with the same number of planning units. Second, we aim to determine if the addition of surrogate-based targets to conservation planning exercises can improve the proportion of genetic variation they secure. We expect that prioritizations generated using amount- and surrogate-based targets will secure more genetic variation than prioritizations generated using only amount-based targets, and that they will secure a similar amount of genetic variation to prioritizations generated using amount- and genetic based targets. Additionally, we expect that prioritizations generated using amount- and surrogate-based targets will share many planning units with prioritizations generated using amount- and genetic-based targets. If these surrogates can improve the intra-specific genetic variation captured in a prioritization, then conservation planners may use these surrogates when genetic data is not available, and in turn deliver more effective protected areas.

@@

### Prioritizations
We used the unreliable representative and adequate prioritization (URAP) formulation of the reserve selection problem [@r465]. This formulation can identify the minimum number of planning units required to preserve both a proportion of the species' range (using amount-based targets) and a proportion of the species' intra-specific variation [using space-based targets and attribute spaces; _sensu_ @r465]. We treated the environmental and geographic surrogate variables as separate attribute spaces. We also treated the adaptive (if detected) and neutral variables as separate attribute spaces for each species. We set the demand points for each species and attribute space using the values associated with the planning units. For every prioritization we generated, we calculated the proportion of the adaptive (if detected) and neutral genetic variation that every prioritization secured to assess its performance. Prioritizations associated with negative values--because they secured only a very small amount of genetic variation--were replaced with zeros to facilitate statistical analysis. We solved all reserve selection problems to within `r (gurobi.params.LST[[MODE]]$MIPGap*100)` % of optimality using the $\texttt{rapr}$ R package [@r465] and $\texttt{Gurobi}$ [@r463].

To address the first aim and determine if the environmental and geographic variables are effective surrogates for adaptive and neutral genetic variation, we generated a collection of single-species prioritizations. For each species, we first generated two sets of prioritizations using increasing environmental and geographic-based targets (separately). Prioritizations were generated using the following surrogate-based targets: `r paste(rapr.params.LST[[MODE]]$surrogacy.analysis$surrogate.target[-length(rapr.params.LST[[MODE]]$surrogacy.analysis$surrogate.target)],collapse=', ')`, and `r rapr.params.LST[[MODE]]$surrogacy.analysis$surrogate.target[length(rapr.params.LST[[MODE]]$surrogacy.analysis$surrogate.target)]`. We then created two corresponding sets of amount-based prioritizations with which to compare the surrogate-based prioritizations. For each species, for each surrogate-based target, we generated `r rapr.params.LST[[MODE]]$surrogacy.analysis$random.replicates` replicate prioritizations by randomly selecting planning units occupied by the species. The number of randomly selected planning units in each prioritization was set to equal the number of planning units in the surrogate-based prioritization.

To address the second aim and determine if the addition of surrogate-based targets can result in more effective prioritizations, we generated a suite of prioritizations using different types of targets. To assess the benefits of surrogate-based targets under the simplest of circumstances, our first scenario involved generating single-species prioritizations for each species. To assess the benefits in a more complex situation, our second scenario involved generating multi-species prioritizations. Since most large-scale prioritizations also explicitly consider opportunity cost to ensure that prioritizations are cost-effective. Our third scenario involved generating multi-species prioritizations that satisfied targets for minimal opportunity cost. For each scenario, we generated three types of prioritizations using (1) only amount-based targets, (2) both amount- and surrogate-based targets, and (3) both amount- and genetic-based targets. The purely amount-based prioritizations served as our baseline--representing the prioritizations typically generated in conservation planning exercises. We used `r (rapr.params.LST[[MODE]]$scenario.analysis$amount.target*100)` % amount-based targets in all prioritizations to ensure an adequate proportion of habitat was secured for each species. Based on the previous analysis, we set surrogate-based targets as `r (rapr.params.LST[[MODE]]$scenario.analysis$surrogate.target*100)` % and genetic-based targets as `r (rapr.params.LST[[MODE]]$scenario.analysis$genetic.target*100)` % to ensure that most of the variation was secured for each species. To accomodate variation in the high number of optimal solutions for the single-species amount-based prioritizations, we generated `r (rapr.params.LST[[MODE]]$scenario.analysis$single.species.amount.replicates)` replicate prioritizations.

## Statistical analysis
We investigated the effectiveness of prioritizations generated using surrogate-based targets using generalized linear mixed-effects models [GLMMs; using a logit link and the $\texttt{lme4}$ R package; @r469]. To address the first aim, we fit a full model to the proportion of genetic variation that the first collection of prioritizations. This full model contained a categorical predictor variable indicating the surrogate-based targets used to generate them, the type of targets used to generate the prioritization (amount or surrogate-based targets), and the type of genetic variation (adaptive or neutral). To address the second aim, we fit a full model to the proportion of genetic variation that second collection of prioritizations secured. This second full model contained a categorical predictor variable indicating the scenario (single-species, multi-species, multi-species with cost), the combination of targets used to generate the prioritizations (amount-based targets, amount and surrogate-based targets, amount and genetic-based targets), and the type of genetic variation measured (adaptive or neutral). Both models contained intercept-level random effects to accommodate variation between species. We also conducted a post-hoc analysis on both models using Tukey contrasts with a Bonferroni correction [using the $\texttt{multcomp}$ R package; @r470].

## Results
<!-- Surrogacy analysis -->
The environmental and geographic-based variables were not effective surrogates for adaptive and genetic variation (Figure 1). Although prioritizations generated using greater surrogate-based targets secured more genetic variation (), they did not secure more genetic variation than prioritizations generated using just amount-based targets (). Moreover, the geographic variables were particularly poor surrogates because they often performed worse than just using amount-based targets on average. Overall, these results suggest that environmental and geographic surrogates are poor predictor for adaptive and neutral genetic variation in the context of conservation planning.
